---
title: "hmclearn package:  Linear Regression Example"
author:  "Samuel Thomas"
date: "``r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hmclearn_linear_regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(hmclearn)
```

Load Scots Race data and display the first few rows of the design matrix $X$.

```{r, echo=TRUE}
library(MASS)

ScotsRaces <- hills

X <- cbind(1, as.matrix(ScotsRaces[, -which(colnames(ScotsRaces) == "time")]))

# add interaction
X <- cbind(X, X[, "dist"] * X[, "climb"])
colnames(X)[ncol(X)] <- "climb_distance"
colnames(X)[1] <- "intercept"

y <- ScotsRaces$time

head(X, 3)
```


# Frequentist linear model

To compare results, we first fit a standard linear model using the frequentist function \textit{lm}

```{r, echo=TRUE}
f <- lm(y ~ X-1)
summary(f)

```

Next, we store the parameter values from the Frequentist fit

```{r, echo=TRUE}
beta.true <- as.numeric(coefficients(f))
sigma2.true <- sigma(f)^2
theta.true <- c(beta.true, sigma2.true)

```

# Prep for HMC

Likelihood function for linear regression

$$
p(y | X, \beta; \sigma^2) = \frac{1}{(2\pi\sigma^2)^{n/2}}\exp{\left(-\frac{1}{2\sigma^2} (y - X\beta)^T(y-X\beta) \right)}
$$

The posterior for $\beta$ and $\sigma^2$

$$
\begin{aligned}
p(\beta, \sigma^2|y; X) &\propto p(y|X, \beta, \sigma^2) p(\beta, \sigma^2) \\
&\propto p(y|X, \beta, \sigma^2)p(\beta)p(\sigma^2)
\end{aligned}
$$

Multivariate Normal prior for $\beta$ and Inverse Gamma for $\sigma^2$

$$
\begin{aligned}
p(\beta) &\propto N(0, BI) \\
p(\sigma^2) &\sim IG(a, b)
\end{aligned}
$$

pdf for $\sigma^2$ prior

$$
p(\sigma^2) = \frac{b^a}{\Gamma(a)} (\sigma^2)^{-a-1} \exp\left(-\frac{b}{\sigma^2} \right)
$$

Log-transform $\sigma^2$ to allow proposals across entire real number line

$$
\begin{aligned}
\gamma &= \log \sigma^2 \\
\sigma^2 &= g^{-1}(\gamma) = e^\gamma \\
p_\gamma(\gamma) &= p_{\sigma^2}(g^{-1}(\gamma))\left\lvert \frac{d\sigma^2}{d\gamma} \right\rvert \\
&= \frac{b^a}{\Gamma(a)}(e^\gamma)^{-a-1} \exp\left(-\frac{b}{\sigma^2} \right) \left\lvert e^\gamma  \right\rvert \\
&= \frac{b^a}{\Gamma(a)}e^{-a\gamma}\exp\left(-\frac{b}{e^\gamma} \right) \\
&\propto e^{-a\gamma}\exp\left(-\frac{b}{e
^\gamma} \right) \\
\log p(\gamma) &\propto -a\gamma - be^{-\gamma}
\end{aligned}
$$

The posterior and log posterior for linear regression, based on $\beta$ and $\gamma$

$$
\begin{aligned}
p(\beta, \gamma|y, X) &\propto p(y|X, \beta, \gamma) p(\beta, \gamma) \\
&\propto p(y|X, \beta, \gamma)p(\beta)p(\gamma) \\
&\propto \frac{1}{(2\pi e^\gamma)^{n/2}}\exp{\left(-\frac{1}{2 e^\gamma} (y - X\beta)^T(y-X\beta) \right)} e^{-a\gamma}\exp\left(-\frac{b}{e^\gamma} \right) \\
&\propto \frac{e^{-\gamma n/2}}{(2\pi)^{n/2}}\exp{\left(-\frac{e^{-\gamma}}{2} (y - X\beta)^T(y-X\beta) \right)} e^{-\frac{1}{2}\beta^T\Sigma_\beta^{-1}\beta} e^{-a\gamma} e^{-b e^{-\gamma}} \\
\log p(\beta, \gamma | y, X) &\propto -\frac{\gamma n}{2} -\frac{e^{-\gamma}}{2} (y - X\beta)^T(y-X\beta) - a\gamma - b e^{-\gamma}  \\
&\propto -\left(\frac{n}{2} + a \right)\gamma  -\frac{e^{-\gamma}}{2} (y - X\beta)^T(y-X\beta)- b e^{-\gamma} - \frac{1}{2}\beta^T \Sigma_\beta^{-1}\beta
\end{aligned}
$$

Next, we derive the gradient of the log posterior


$$
\begin{aligned}
\log p(\beta, \gamma | y, X) &\propto -\frac{\gamma n}{2} -\frac{e^{-\gamma}}{2} (y - X\beta)^T(y-X\beta) - a\gamma - b e^{-\gamma}  \\
&\propto -\left(\frac{n}{2} + a \right)\gamma  -\frac{e^{-\gamma}}{2} (y - X\beta)^T(y-X\beta)- b e^{-\gamma} \\
\frac{\partial}{\partial \beta} \log p(\beta, \gamma | y, X) &\propto -\frac{e^{-\gamma}}{2}(-2)X^T(y-X\beta) - \Sigma_\beta^{-1}\beta\\
&\propto e^{-\gamma} X^T(y-X\beta) \\
\frac{\partial}{\partial\gamma}\log p(\beta, \gamma | y, X) &\propto -\left(\frac{n}{2} + a \right) + \frac{e^{-\gamma}}{2} (y - X\beta)^T(y-X\beta)+ b e^{-\gamma}
\end{aligned}
$$

Fit model using hmc

```{r, echo=TRUE}
N <- 1e5

# eps_vals <- c(1e-3, 2e-3, rep(2e-6, 2), 2e-4)
eps_vals <- c(1e-1, 1e-3, 1e-5, 4e-7, 1e-3)

set.seed(412)
t1.hmc <- Sys.time()
 res <- hmc(N=N, theta.init = rep(0, 5), 
                    epsilon = eps_vals, L = 20, 
                    logPOSTERIOR = linear_posterior, 
                    glogPOSTERIOR = g_linear_posterior, 
                    param=list(y = y, X=X), 
                    varnames = c(colnames(X), "log_sigma2"),
                    parallel=TRUE, chains=4)
t2.hmc <- Sys.time()
t2.hmc - t1.hmc

res$accept/N
```

```{r, echo=TRUE}
summary(res, burnin=5000)
```


```{r, echo=TRUE}
diagplots(res, actual.mu=truevals, burnin=3000)


```
