---
title: "hmclearn package:  Logistic Mixed Effects Regression Example"
author: "Samuel Thomas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{logistic_mixed_effects_hmclearn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette demonstrates fitting a Logistic mixed effects regression model via Hamiltonian Monte Carlo (HMC) using the **hmclearn** package. 

For a mixed effects model with binary response, we let

$$
p = Pr(Y = 1 | X, Z) = [1 + e^{-X\beta-Zu}]^{-1}
$$
or

$$
\begin{aligned}
\text{logit}[P(y = 1 | u)] &= X\beta + Zu \\
u &\sim N(0, G)
\end{aligned}
$$


```{r setup}
library(hmclearn)
```


# Logistic Mixed Effects Model Example Data

The user must define provide the design matrix directly for use in **hmclearn**.  Our first step is to load the data and store the fixed effect design matrix $X$, random effects design matrix $Z$, and dependent variable vector $y$. 

We load drug Contraception data (Bates, et. al. 2014) and create the design matrices $X$ and $Z$ and dependent vector $y$.  For this model, the random effects design matrix $Z$ is specified for a random intercept model.  

```{r, echo=TRUE}
Contraception <- mlmRev::Contraception

Contraception$liv2 <- ifelse(Contraception$livch == "0", 0, 1)

##########
# block diagonal
Zi.lst <- split(rep(1, nrow(Contraception)), Contraception$district)
Zi.lst <- lapply(Zi.lst, as.matrix)
Z <- Matrix::bdiag(Zi.lst)
Z <- as.matrix(Z)

urban <- ifelse(Contraception$urban == "Y", 1, 0)

X <- cbind(1, Contraception$age, Contraception$age^2, urban, Contraception$liv2)
colnames(X) <- c("int", "age", "age_sq", "urban", "liv2")
y <- ifelse(Contraception$use == "Y", 1, 0)

```


## QR decomposition of design matrix

To facilitate a more efficient fitting of the model, we apply QR decomposition to the fixed effects design matrix $X$.  

Let $\theta = R^*\beta$ (below).  The HMC estimates $\theta$, from which we can use the deterministic formula to determine $\beta$

$$
\begin{aligned}
X &= Q^* R^* \\
Q^* &= Q \cdot \sqrt{n-1} \\
R^* &= \frac{1}{\sqrt{n-1}}R \\
X\beta &= Q^* R^* \beta \\
\beta &= R^{*^{-1}}\theta
\end{aligned}
$$

```{r, echo=TRUE}
xqr <- qr(X)
Q <- qr.Q(xqr)
R <- qr.R(xqr)

n <- nrow(X)
X2 <- Q * sqrt(n-1)
Rstar <- R / sqrt(n-1)
Rstar_inv <- solve(Rstar)
colnames(X2) <- c("int", "age", "age_sq", "urban", "liv2")

# new intercept from QR decomposition
diagval <- X2[1,1]


##########
# block diagonal
Zi.lst <- split(rep(diagval, nrow(Contraception)), Contraception$district)
Zi.lst <- lapply(Zi.lst, as.matrix)
Z2 <- Matrix::bdiag(Zi.lst)
Z2 <- as.matrix(Z2)

```


## Likelihood Derivation

First, we derive the likelihood for our logistic mixed effects regression model.

$$
\begin{aligned}
p(y | X, Z, \beta, u) &= \prod_{i=1}^n\prod_{j=1}^m \left(\frac{1}{1 + e^{-X_{i}\beta - Z_{ij}u_i}}\right)^{y_{ij}} \left(\frac{e^{-X_i\beta - Z_{ij}u_i}}{1 + e^{-X_{i}\beta - Z_{ij}u_i}}\right)^{1-y_{ij}} \\
\log p(y|X,Z,\beta,u) &= \sum_{i=1}^n\sum_{j=1}^m -y_{ij}\log (1 + e^{-X_i\beta - Z_{ij}u_i}) + (1 - y_{ij})\log e^{-X_i\beta - Z_{ij}u_i} \\
&- (1-y_{ij})\log(1 + e^{-X_i\beta - Z_{ij}u_i}) \\
&= \sum_{i=1}^n \sum_{j=1}^m -y_{ij}\log (1 + e^{-X_i\beta - Z_{ij}u_i}) - (1-y_{ij})(X_{ij}\beta + Z_{ij}u_i) \\
&- (1-y_{ij})\log(1 + e^{-X_i\beta - Z_{ij}u_i}) \\
&= \sum_{i=1}^n\sum_{j=1}^m -y_{ij}\log(1 + e^{-X_i\beta-Z_{ij}u_i}) - (X_{ij}\beta + Z_{ij}u_i - \log(1+e^{-X_i\beta-Z_{ij}u_i})) \\
&+ y_{ij}(X_{ij}\beta + Z_{ij}u_i + \log(1 + e^{-X_i\beta - Z_{ij}u_i})) \\
&=\sum_{i=1}^n \sum_{j=1}^m -(1 - y_{ij})(X_{i}\beta + Z_{ij}u_i) - \log(1 + e^{-X_{i}\beta - Z_{ij}u_i}) \\
&= -(1 - y)^T(X\beta + Zu) - \log(1 + e^{-\sum_{i=1}^n \sum_{j=1}^m (X_i\beta + Z_{ij}u_i)})
\end{aligned}
$$

# Priors

Priors are needed for our parameters $\beta$ and $G$ to formulate the posterior.  

We set a multivariate Normal prior for $\beta$

$$
\begin{aligned}
\beta &\sim N(0, \Sigma_\beta) \\
&\sim N(0, \sigma_\beta^2I)
\end{aligned}
$$


with pdf

$$
\begin{aligned}
p(\beta) &= \frac{1}{\sqrt{\lvert 2\pi \Sigma_\beta \rvert }}e^{-\frac{1}{2}\beta^T \Sigma_\beta^{-1}\beta} \\
\log p(\beta) &= -\frac{1}{2}\log(2\pi \lvert \Sigma_\beta \rvert) - \frac{1}{2}\beta^T \Sigma_\beta^{-1} \beta \\
&\propto -\frac{1}{2}\log \lvert\Sigma_\beta\rvert - \frac{1}{2}\beta^T \Sigma_\beta^{-1} \beta.
\end{aligned}
$$


# Parameterization for $G$

The parameterization approach for this model uses a strategy recommended by Betancourt, Girolami (2013) to facilitate more efficient sampling in HMC.  

Further, the uniform parameterization of the variance parameters is replaced by a half-t family of distributions per Gelman (2006), Prior distributions for Variance parameters in hierarchical models.  This parameterization is well-behaved around 0, in contrast to inverse gamma, and provides flexibility for more informed priors than a uniform distribution.

We select a parameterization of $G$ such that the likelihood and its gradient can be derived for HMC.  To this end, we uses LDL decomposition of $G$ to form a flexibile parameterization that can easily handle restrictions (Chan, Jelizkov 2009).  

$$
\begin{aligned}
u &\sim N(0, G)  \\
G &= L D L^T \\
&= L D^{1/2} D^{1/2} L^T \\
\end{aligned}
$$

Let $\lambda_k$ where $k = 1, ... p$ denote the diagonal entries of $D^{1/2}$ and let $a_{kj}$ where $1 \leq j < k \leq p$ denote free elements of lower unitrangular matrix $L$

$$
D^{1/2} := 
\begin{pmatrix}
\lambda_1 & 0 & ... & 0 \\
0 & \lambda_2 & 0 ... & 0 \\
... & ... & ... & ... \\
0 & 0 & ... & \lambda_p
\end{pmatrix}, 
L :=
\begin{pmatrix}
1 & 0 & 0 & ... & 0 \\
a_{21} & 1 & 0 & ... & 0 \\
a_{31} & a_{32} & 1 & ... & ... \\
... & ... & ... & ... & ... \\
a_{p1} & a_{p2} & ... & ... & 1 \\
\end{pmatrix}
$$

Also define $\lambda := (\lambda_1, ..., \lambda_p)^T$ and $a_k := (a_{k1}, ..., a_{k, k-1})^T$ and $a := (a_2^T, ..., a_p^T)^T$

Consider priors where $k = 1...p$.  The prior for $\lambda_k$ is half-t per Gelman (2006).  

$$
\begin{aligned}
p(\lambda_k) &\sim \left(1 + \frac{1}{\nu}\left(\frac{\lambda_k}{A} \right)^2 \right)^{-(\nu+1)/2} \\
a|\lambda &\sim N(a_0, A_0)
\end{aligned}
$$

The hyperparameter $a_0$ does not need to be zero, and $A_0$ can be correlated and may depend on $\lambda$.  In this model, we define $a$ independent of $\lambda$. 

Per Betancourt, Girolami (2013), we re-parameterize $u$ using a standard normal parameterization we define as $\tau = (\tau_1, ..., \tau_q)$.  Here, $u$ is a deterministic function of $G$ and $\tau$.


$$
\begin{aligned}
\tau &\sim N(0, I_q) \\
u &:= L D^{1/2} \tau \\
&\sim N(0, LD^{1/2} I (L D^{1/2})^T) \\
&\sim N(0, L D^{1/2} D^{1/2} L^T) \\
&\sim N(0, G)
\end{aligned}
$$

The distribution of $u$ therefore does not change with this parameterization.  The intent of our re-parameterization is to allow $G$ and $\tau$ to be largely independent in the MCMC sampling.   

## Log posterior derivation

Now that we have the log likelihood and priors specified, we can derive the log posterior.  

$$
\begin{aligned}
p(\beta, u, G | y, X, Z) &\propto p(y | \beta, u, G)  p(\beta, u, G) \\
&\propto p(y|\beta, u, G) p(\beta) p(u|G) p(G) \\
\log p(\beta, u, G | y, X, Z) &\propto \log p(y|\beta, u, G) + \log p(\beta) + \log p(u|G) + \log p(G)
\end{aligned}
$$

The log posterior is the sum of the log likelihood and the log of the prior for $\beta$.  

We have hyperpriors $\lambda$ and $a$ for $G$.  Also, we use a half-t prior for variance parameters per Gelman (2006). 

$$
\begin{aligned}
p(\lambda_k) &\sim \left(1 + \frac{1}{\nu}\left(\frac{\lambda_k}{A} \right)^2 \right)^{-(\nu+1)/2} \\
p_{\xi_k}(\xi_k) &= p_{\lambda_k}(g^{-1}(\xi_k)) \left\lvert \frac{d\lambda_k}{d\xi_k}  \right\rvert \\
&= p_{\lambda_k} (e^{\xi_k})\lvert e^{\xi_k}\rvert \\
&\propto  \left(1 + \frac{1}{\nu}\left(\frac{e^{\xi_k}}{A} \right)^2 \right)^{-(\nu+1)/2} e^{\xi_k}\\
&\propto \left(1 + \frac{1}{\nu}\left(\frac{e^{2\xi_k}}{A^2} \right) \right)^{-(\nu+1)/2} e^{\xi_k}\\
\log p(\xi_k) &\propto -\frac{\nu+1}{2}\log\left(1 + \frac{1}{\nu}\left(\frac{e^{2\xi_k}}{A^2} \right) \right) + \xi_k \\
\frac{\partial}{\partial\xi_k}\log p(\xi_k) &\propto -\frac{\nu+1}{2}\frac{1}{1 + \frac{1}{\nu}\left( \frac{e^{2\xi_k}}{A^2} \right)}\frac{2e^{2\xi_k}}{\nu A^2} + 1 \\
&\propto -(\nu+1)\frac{1}{1 + \nu A^2 e^{-2\xi_k}} + 1
\end{aligned}
$$

We assign a relatively uniformative prior for $a$

$$
\begin{aligned}
a_k &\sim N(0, A) \\
p(a_k) &\propto \lvert A  \rvert^{-1/2} e^{-\frac{1}{2} a_k^T A^{-1} a_k} \\
\log p(a_k) &\propto -\frac{1}{2}a_k^T A^{-1} a_k
\end{aligned}
$$

The gradient with respect to $a$:

Note that the following are equivalent

$$
\begin{aligned}
L D^{1/2}\tau = D^{1/2} \tau + \widetilde{T} a
\end{aligned}
$$

Therefore, the portion of the log likelihood dependent on $a$ becomes

$$
\begin{aligned}
l(a, ..) &\propto -(1 - y)^T(X\beta + ZLD^{1/2}\tau) - \log(1 + e^{-\sum_{i=1}^n \sum_{j=1}^m (X_i\beta + Z_{ij}LD^{1/2}\tau_j)}) \\
&-\frac{1}{2}\log \lvert\Sigma_\beta\rvert - \frac{1}{2}\beta^T \Sigma_\beta^{-1} \beta \\
&-\sum_{k=1}^q \left(\frac{\nu_{\lambda_k}+1}{2}\log\left(1 + \frac{1}{\nu_{\lambda_k}} 
\frac{e^{2\xi_k}}{A_{\lambda_k}^2}\right)+ \xi_k\right)\\
&-\frac{1}{2} a^T A^{-1} a -\frac{1}{2} \tau^T \tau\\
&\propto -(1 - y)^T Z\widetilde{T} - \left(\frac{e^{-(X\beta + ZLD^{1/2}\tau)}}{1 + e^{-(X\beta + ZLD^{1/2}\tau)}}\right)^T  Z\widetilde{T} - A^{-1}a
\end{aligned}
$$


Finally, we write the full log posterior

Note: we use $A_\xi$ and $A_a$ to distinguish between hyperpriors for $\xi$ and for $a$

$$
\begin{aligned}
\log p(\beta, u, G | y, X, Z) &\propto \log p(y|\beta, u, G) + \log p(\beta) + \log p(u|G) + \log p(G) \\
\log p(\beta, \gamma, \tau, \xi_1...\xi_q,a | y, X, Z) &\propto \log p(y | \beta, \tau, \gamma, \xi, a) + \log p(\beta) + \log p(\gamma)+ \log p(\tau) + \log p(\xi) + \log p(a) \\
&\propto  -(1 - y)^T(X\beta + Zu) - \log(1 + e^{-\sum_{i=1}^n \sum_{j=1}^m (X_i\beta + Z_{ij}u_i)}) \\
&-\frac{1}{2}\log \lvert\Sigma_\beta\rvert - \frac{1}{2}\beta^T \Sigma_\beta^{-1} \beta \\
&-\sum_{k=1}^q \left(\frac{\nu_{\lambda_k}+1}{2}\log\left(1 + \frac{1}{\nu_{\lambda_k}} 
\frac{e^{2\xi_k}}{A_{\lambda_k}^2}\right)+ \xi_k\right)\\
&-\frac{1}{2} a^T A^{-1} a -\frac{1}{2} \tau^T \tau\\
&\propto  -(1 - y)^T(X\beta + ZLD^{1/2}\tau) - \log(1 + e^{-\sum_{i=1}^n \sum_{j=1}^m (X_i\beta + Z_{ij}LD^{1/2}\tau_j)}) \\
&-\frac{1}{2}\log \lvert\Sigma_\beta\rvert - \frac{1}{2}\beta^T \Sigma_\beta^{-1} \beta \\
&-\sum_{k=1}^q \left(\frac{\nu_{\lambda_k}+1}{2}\log\left(1 + \frac{1}{\nu_{\lambda_k}} 
\frac{e^{2\xi_k}}{A_{\lambda_k}^2}\right)+ \xi_k\right)\\
&-\frac{1}{2} a^T A^{-1} a -\frac{1}{2} \tau^T \tau\\
\end{aligned}
$$


The gradient of the log posterior must be derived for the leapfrog function

$$
\begin{aligned}
\log p(\beta, \gamma, \tau, \xi_1...\xi_q,a | y, X, Z) &\propto  -(1 - y)^T(X\beta + ZLD^{1/2}\tau) - \log(1 + e^{-\sum_{i=1}^n \sum_{j=1}^m (X_i\beta + Z_{ij}LD^{1/2}\tau_j)}) \\
&-\frac{1}{2}\log \lvert\Sigma_\beta\rvert - \frac{1}{2}\beta^T \Sigma_\beta^{-1} \beta \\
&-\sum_{k=1}^q \left(\frac{\nu_{\lambda_k}+1}{2}\log\left(1 + \frac{1}{\nu_{\lambda_k}} 
\frac{e^{2\xi_k}}{A_{\lambda_k}^2}\right)+ \xi_k\right)\\
&-\frac{1}{2} a^T A^{-1} a -\frac{1}{2} \tau^T \tau\\
\frac{\partial l}{\partial\beta} &\propto -(1-y)^T X + \left(\frac{e^{-(X\beta + Zu)}}{1 + e^{-(X\beta + Zu)}}\right)^T X - \Sigma_\beta^{-1}\beta\\
\frac{\partial l}{\partial \tau} &\propto -(1-y)^T ZLD^{1/2} + \left(\frac{e^{-(X\beta + Zu)}}{1 + e^{-(X\beta + Zu)}}\right)^T ZLD^{1/2} - \tau\\
\frac{\partial l}{\partial \xi_k} &= -(1-y)^T ZLJ^{kk}\tau + \left(\frac{e^{-(X\beta + Zu)}}{1 + e^{-(X\beta + Zu)}}\right)^T ZLJ^{kk}\tau- (\nu_{\lambda_k}+1)\frac{1}{1 + \nu_{\lambda_k}A_{\lambda_k}^2 e^{-2\xi_k}}+1 \\
\frac{\partial l}{\partial a} &= -(1 - y)^T Z\widetilde{T} - \left(\frac{e^{-(X\beta + ZLD^{1/2}\tau)}}{1 + e^{-(X\beta + ZLD^{1/2}\tau)}}\right)^T  Z\widetilde{T} - A^{-1}a
\end{aligned}
$$

Run HMC for logistic regression model using QR decomposed design matrices

```{r, echo=TRUE}
N <- 2e3

set.seed(412)
initvals<- c(rep(0, 5), # fixed effects
                rnorm(60, mean=0, sd=0.1), # random intercepts
                0) # variance of random intercepts


vnames <- c(colnames(X), 
            paste0("tau_int", 1:60), 
            "xi1")

epsvals <- c(5e-2, rep(1e-2, 4), rep(5e-2, 61))

t1.hmc <- Sys.time()
f_hmc <- hmc(N = N, theta.init = initvals,
             epsilon = epsvals, L = 10,
             logPOSTERIOR = glmm_bin_posterior,
             glogPOSTERIOR = g_glmm_bin_posterior,
             varnames = vnames,
             parallel=TRUE, chains=2,
             param=list(y = y, X=X2, Z=Z2, n=60, nrandom=1, sig2beta=5,
                        nuxi=1, Axi=25)  )


t2.hmc <- Sys.time()
t2.hmc - t1.hmc
```

The acceptance ratio for each of the HMC chains is sufficiently high for an efficient simulation.  

```{r, echo=TRUE}
f_hmc$accept/N
```

Trace plots provide a visual indication of stationarity.  These plots indicate that the MCMC chains are reasonably stationary.  

```{r, echo=TRUE}
mcmc_trace(f_hmc, burnin=1000, pars=colnames(X))
```

Since we used QR decomposition to transform $\beta$ prior to fitting via HMC, we need to reverse the transformation to obtain the original parameter scale.  

```{r, echo=TRUE}
# restore beta from Rstar in QR decomposition
calc_beta <- function(theta_param, Rstarinv) {
  as.numeric(Rstarinv %*% theta_param)
}

# reverse qr decomposition
f_hmc2 <- f_hmc
f_hmc2$thetaCombined <- lapply(f_hmc$thetaCombined, function(xx) {
  xx[, 1:5] <- t(apply(xx[, 1:5], 1, calc_beta, Rstarinv=Rstar_inv))
  xx
})

```

The revised summary shows the posterior distribution after transformation back to the original scale.  

```{r, echo=TRUE}
summary(f_hmc2)
```

We create trace plots on the transformed simulation data. 

```{r, echo=TRUE, fig.width=6, fig.height=4}
mcmc_trace(f_hmc2, burnin=1000, pars=colnames(X))
```

# Frequentist model

To compare results, we first fit a logistic mixed effects model using the frequentist package **lme4** (Bates et. al. 2015).  

```{r, echo=TRUE}
library(lme4)

f <- glmer(use ~ age + I(age^2) + urban + liv2 + (1 | district), 
           data=Contraception, family=binomial, 
           control=glmerControl(optCtrl=list(maxfun=20000)))
summary(f)
```

```{r, echo=TRUE}

freqvals <- c(as.numeric(fixef(f)), 
              as.numeric(ranef(f)$district[, 1]), 
              log(sqrt(as.numeric(VarCorr(f)[1]))))
```


Histograms of the posterior distribution show that Bayesian parameter estimates align with frequentist estimates.  The *cols* parameter specifies the parameters to be displayed in *diagplots*, based on the order provided to the *hmc* function. 

```{r, echo=TRUE, fig.width=6, fig.height=4}
diagplots(f_hmc2, burnin=1000, 
          actual.mu=c(freqvals[1:ncol(X)], freqvals[length(freqvals)]), 
          cols=c(1:ncol(X), length(initvals)))

```


# Source

Steele, F., Diamond, I. And Amin, S. (1996). Immunization uptake in rural Bangladesh: a multilevel analysis. *Journal of the Royal Statistical Society*, Series A (159): 289-299.  

\url{http://www.bristol.ac.uk/cmm/learning/mmsoftware/data-rev.html}

# References

Bates, D., M\"{a}chler, M., Bolker, B., & Walker, S. (2015). Fitting linear mixed-effects models using lme4. *Journal of Statistical Software* 67(1)

Bates, D., M\"{a}chler, M., & Bolker, B. (2014). mlmRev: Examples from multilevel modelling software review. *R package version, 1*.

Agresti, A. (2015). *Foundations of linear and generalized linear models*. John Wiley & Sons.  ISBN: 978-1-118-73003-4
